---
title: "epiCo - Demographic module"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{epiCo - Demographic module}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

**epiCo**'s demographic module is a tool for demographic descriptive
analysis and risk assessment of epidemiological events in Colombia.
Based on linelist data provided by the Colombian National Surveillance System
[SIVIGILA](https://www.ins.gov.co/Direcciones/Vigilancia/Paginas/SIVIGILA.aspx)
and demographic data from the Colombian National Administrative Department of 
Statistics 
[DANE](https://microdatos.dane.gov.co/index.php/catalog/central/about).

The module allows you to:

-   Consult and visualize the population pyramid of a municipality,
    department, or the country for a year of interest.
-   Consult the definitions of demographic variables as ethnicities,
    special populations groups, and occupational labels.
-   Estimate the distribution of occupations reported in the linelist
    according to the [ISCO 88
    Standard](https://www.ilo.org/public/english/bureau/stat/isco/isco88/).
-   Estimate the incidence rate of a municipality, department, or the country. 
-   Calculate and visualize a relative age risk by normalizing the
    distribution of cases by the specific age structure of the location
    of interest.
    
In the following vignette you will learn how to:

1. Navigate the Codification of the Political Administrative Division of Colombia (DIVIPOLA).
2. Consult, visualize and interpret Colombian population pyramids at different
population scales.
3. Interpret the demographic variables reported by the SIVIGILA (as ethnicities,
special population groups, and occupatioanl labels).
4. Understand the epidemiological data example.
5. Estimate weekly and monthly global incidence rates for a municipality, department or the country.
6. Integrate age distributions of cases with population pyramids to obtain an age
risk assessment for a disease.

## 1. Navigating the Codification of the Political Administrative Division of Colombia (DIVIPOLA)

DIVIPOLA is a standardized nomenclature, designed by the DANE, for the identification of territorial entities (departments, districts and municipalities), non-municipalized areas and populated centers, by assigning a unique numerical code to each of these territorial units.

Colombia has:

- 32 Departments (Administrative Division Level 1)
- 1102 Municipalities (Administrative Division Level 2)
- 1 Island
- 18 Non-municipalized areas
- 6678 Populated centers

Two digits are used for the codification of Departments, and five digits are used for the codification of municipalities (the first two being the the department where they are located).

**epiCo** provides the complete list of departments and municipalities codes through a built-in dataset.

```{r}
library(epiCo)
library(incidence)

data("divipola_table")
```
## 2. Population pyramids

**epiCo** provides built-in dataset with the population projections of Colombia at national, department, and municipality level (provided by the DANE).

You can perform queries to these data by using the 'population_pyramid' function, providing the DIVIPOLA code of the territory of interest and the year to consult.

```{r}
ibague_code <- 73001 # DIVIPOLA code for the city of Ibagué
year <- 2019 # Year to consult
ibague_pyramid_2019 <- population_pyramid(ibague_code, year) # Population 
# pyramid (dataframe) for the city of Ibagué in the year 2019 dissagregated 
# by sex
head(ibague_pyramid_2019)
```

Definition of age ranges, and plotting are also provided for both: total number of individuals or proportion of individuals

```{r}
ibague_code <- 73001 # DIVIPOLA code for the city of Ibagué
year <- 2016 # Year to consult
age_range <- 5 # Age range or window
ibague_pyramid_2016 <- population_pyramid(ibague_code, year, range = age_range, 
                                          gender=TRUE, total = FALSE, 
                                          plot = TRUE)
```


## 3. Demographic variables

Events of epidemiological relevance are reported to the SIVIGILA using an official notification form (see, [link](https://www.ins.gov.co/buscador-eventos/Paginas/Fichas-y-Protocolos.aspx)).

**epiCo** provides function to consult the dictionaries to the etnicity cathegories, special population groups, and occupation codification used by the SIVIGILA.

```{r}
demog_data <- data.frame(
  id = c(0001, 002, 003, 004, 005),
  ethnicity_label = c(3, 4, 1, 2, 5),
  occupation_label = c(6111, 3221, 5113, 5133, 6111)
)

describe_ethnicity(demog_data$ethnicity_label) # Description of ethnicities 
# reported in the consulted dataset
describe_occupation(isco_codes = demog_data$occupation_label, 
                    output_level = "unit") # Description of the occupations
# reported in the consulted dataset
```
## 4. Epidemiological data

**epiCo** is a tool that produces analyses based on epidemiological data extracted from SIVIGILA or provided by the user. In the data folder is the epi_data file, which shows an example of the structure used by the package, which is the same as the one reported by SIVIGILA. This file contains the cases of all the municipalities in Tolima for the years 2015–2021.

The following analyses use the dengue cases reported in Tolima in 2019.

```{r}
data("epi_data")
data_event <- epi_data

data_tolima <- data_event[lubridate::year(data_event$fec_not) == 2019, ]
head(data_tolima)
```

## 5. Estimation of incidence rates

The incidence rate feature of **epiCo** requires the [incidence2](https://github.com/reconverse/incidence2) package to produce a modified incidence object where, instead of a count vector (or matrix). It transforms the object to provide a rate element accounting for the number of cases in the time period divided by the total number of inhabitants in the region and year.

**epiCo** uses the DANE population projections as denominators; therefore, it is necessary to provide the administration level at which incidences are calculated.

```{r}
incidence_object <- incidence(dates = data_tolima$fec_not, 
                              groups = data_tolima$cod_mun_o, 
                              interval = "1 epiweek")
incidence_rate_object <- incidence_rate(incidence_object, level = 2) 
head(incidence_rate_object$counts)
```

If groups in the incidence object are not within the DIVIPOLA coding for municipalities (level 2) or departments (level 1), or a national estimation is intended (level 0), the function will not be able to estimate an incidence rate.

## 6. Estimation of risk by age group

Normalization of data is a key aspect on epidemiology. **epiCo** allows to take an age distribution of cases and normalize the epidemiological data with the age structure of a population. This normalization allows to estimate the age risk of a disease according to the age structure of the general population in a municipality, department, or the country in a certain year.

```{r}
data_ibague <- data_tolima[data_tolima$cod_mun_o == 73001, ]

age_risk_data <- age_risk(age = data_ibague$edad, 
                          population_pyramid = ibague_pyramid_2019, 
                          gender = TRUE, plot = TRUE)
```

