---
title: "epiCo - Demographic module"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{epiCo Demographic Analyses}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

**epiCo**'s demographic module is a tool for demographic descriptive
analysis and risk assessment of epidemiological events in Colombia.
Based on linelist data provided by the Colombian National Surveillance System
[SIVIGILA](https://www.ins.gov.co/Direcciones/Vigilancia/Paginas/SIVIGILA.aspx)
and demographic data from the Colombian National Administrative Department of 
Statistics 
[DANE](https://microdatos.dane.gov.co/index.php/catalog/central/about).

The module allows you to:

-   Consult and visualize the population pyramid of a municipality,
    department, or the country for a year of interest
-   Consult the definitions of demographic variables as ethnicities,
    special populations groups, and occupational labels
-   Calculate and visualize a relative age risk by normalizing the
    distribution of cases by the specific age structure of the location
    of interest
-   Visualize the distribution of occupations reported in the linelist
    according to the [ISCO 88
    Standard](https://www.ilo.org/public/english/bureau/stat/isco/isco88/)

    
In the following vignette you will learn how to:

1. Navigate the Administrative Political Division Coding of Colombia (DIVIPOLA)
2. Consult, visualize and interpret Colombian population pyramids at different
population scales
3. Interpret the demographic variables reported by the SIVIGILA (as ethnicities,
special population groups, and occupatioanl labels)
4. Obtain official historic linelist data from SIVIGILA using
Epiverse-TRACE package [sivirep](https://github.com/epiverse-trace/sivirep)
5. Integrate age distributions of cases with population pyramids to obtain an age
risk assessment for a disease
6. Generate a treemap of reported occupations to estimate an exposure risk
assessment for a disease

```{r setup}
library(epiCo)
```
